### 2022-02-26 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_networkary_classifiers@ff3a0ed160a7f03f2cd8bc675822c2f3b0bb1bad 🚀

Informedness (Youden, 1950) (also known as bookmaker informedness or the
True Skill Statistic) is \(\text{TPR}+\text{TNR}-1\), where
\(\text{TPR}= tp/(tp+fn)\) and \(\text{TNR} = tn/(tn+fp)\); informedness
can be used to find the optimal cutpoint in thresholding analyses
(Schisterman et al., 2005). The formula for informedness is

@@

Finally, \(F_1\) is the harmonic mean of precision (the chance that a
positive even was correctly classified) and sensitivity (the ability to
correctly classify positive events), and is defined as

@@

In this section, we will assume a network of connectance \(\rho\),
\emph{i.e.} having \(\rho S^2\) interactions (where \(S\) is the species
richness), and \((1-\rho) S^2\) non-interactions. Therefore, the vector
describing the \emph{true} state of the network (assumed to be an
unweighted, directed network) is a column vector
\(\mathbf{o}^T = [\rho, (1-\rho)]\) (we can safely drop the \(S^2\)
terms, as we will work on the confusion matrix, which ends up expressing
\emph{relative} values). We will apply skill and bias to this matrix,
and measure how a selection of performance metrics respond to changes in
these values, in order to assess their suitability for model evaluation.

@@

Note that when \(s=0\), \(\text{Tr}(\mathbf{M}) = 0\) (the classifier is
\emph{always} wrong), when \(s=0.5\), the classifier is no-skill and
guesses at random, and when \(s=1\), the classifier is perfect.

@@

In all further simulations, the confusion matrix \(\mathbf{C}\) is
transformed so that it sums to 1.

@@

In this section, we will change the values of \(b\), \(s\), and
\(\rho\), and report how the main measures discussed in the introduction
(MCC, \(F_1\), \(\kappa\), and informedness) are responding to issues
with the classifier. Before we do so, it is important to explain why we
will not focus on accuracy too much. Accuracy is the number of correct
predictions (\(\text{Tr}(\mathbf{C})\)) divided by the sum of the
confusion matrix. For a no-skill, no-bias classifier, accuracy is equal
to \(\rho^2 + (1-\rho)^2\); for \(\rho = 0.05\), this is
\(\approx 0.90\), and for \(\rho = 0.01\), this is equal to
\(\approx 0.98\). In other words, the values of accuracy are expected to
be so high that they are not really informatived (this is simply
explained by the fact that for \(\rho\) small,
\(\rho^2 \ll (1-\rho)^2\)). More concerning is the fact that introducing
bias changes the response of accuracy in unexpected ways. Assuming a
no-skill classifier, the numerator of accuracy becomes
\(b\rho^2 + (1-b)(1-\rho)^2\), which increases when \(b\) is low, which
specifically means that at equal skill, a classifier that under-predicts
interactions will have higher accuracy than an un-biased classifier.
These issues are absent from balanced accuracy, but should nevertheless
lead us to not report accuracy as the primary measure of network
prediction success; moving forward, we will focus on other measures.

In order to examine how MCC, \(F_1\), \(\kappa\), and informedness
change w.r.t. the imbalance, skill, and bias, we performed a grid
exploration of the values of \(\text{logit}(s)\) and \(\text{logit}(b)\)
linearly from \(-10\) to \(10\); \(\text{logit}(x) = -10\) means that
\(x\) is essentially 0, and \(\text{logit}(x) = 10\) means it is
essentially 1 -- this choice was motivated by the fact that most
responses are non-linear with regards to bias and skill. The values or
\(\rho\) were taken linearly in \(]0, 0.5]\), which is within the range
of usually observed connectance values for empirical species interaction
networks. Note that at this point, there is no food web model to speak
of; rather, the confusion matrix we discuss can be obtained for any
classification task. Based on the previous discussion, the desirable
properties for a measure of classifier success should be: an increase
with classifier skill, especially at low bias; a hump-shaped response to
bias, especially at high skill, and ideally center around
\(\text{logit}(b)=0\); an increase with prevalence up until
equiprevalence is reached.

@@

In the following section, we will generate random bipartite networks
(this works without loss of generality on unipartite networks), and
train four binary classifiers (as well as an ensemble model using the
sum of ranged outputs from the component models) on 30\% of the
interaction data. In practice, testing usually uses 70\% of the total
data; for ecological networks, where interactions are sparse \emph{and}
the number of species is low, this may not be the best solution, as the
testing set becomes constrained not by the \emph{proportion} of
interactions, but by their \emph{number}. Preliminary experiments using
different splits revealed no qualitative change in the results. Networks
are generated by picking a random infectiousness trait \(v_i\) for 100
species (from a beta distribution \(B(\alpha=6,\beta=8)\) distribution),
and a resistance trait \(h_j\) for 100 species (from
\(B(\alpha=2,\beta=8)\) distribution). There is an interaction between
\(i\) and \(j\) when \(v_i-\xi/2 \le h_j \le v_i+\xi/2\), where \(\xi\)
is a constant regulating the connectance of the network (there is an
almost 1:1 relationship between \(\xi\) and connectance), and varies
uniformly in \([0.05, 0.35]\). This model gives fully interval networks
that are close analogues to the bacteria--phage model of Weitz et al.
(2005), with both a modular structure and a non-uniform degree
distribution. This dataset is easy for almost any algorithm to learn:
when trained with features \([v_i, h_j, \text{abs}(v_i, h_j)] ^T\) to
predict the interactions between \(i\) and \(j\), all four models
presented below were able to reach almost perfect predictions all the
time (data not presented here) -- this is in part because the rule
(there is maximum value of the distance between traits for which there
is an interaction) is fixed for all interactions. In order to make the
problem more difficult to solve, we use \([v_i, h_j]\) as a feature
vector (\emph{i.e.} the traits on which the models are trained), and
therefore the models will have to uncover that the rule for interaction
is \(\text{abs}(v_i, h_j) \le \xi\). The models therefore all have the
form

\[
\begin{bmatrix}
           x_{1} \\
           x_{2} \\
           \vdots \\
           x_{m}
         \end{bmatrix}
\]

@@

>>>>

Informedness (Youden, 1950) (also known as bookmaker informedness or the
True Skill Statistic) is \(\text{TPR}+\text{TNR}-1\), where
\(\text{TPR}= tp/(tp+fn)\) and \(\text{TNR} = tn/(tn+fp)\). Informedness
can be used to find the optimal cutpoint in thresholding analyses
(Schisterman et al., 2005); indeed, the maximal informedness corresponds
to the point on the ROC curve that is closest to the perfect classifier
point. The formula for informedness is

@@

Finally, \(F_1\) is the harmonic mean of precision (the chance that
interaction was correctly detected as such) and sensitivity (the ability
to correctly classify interactions), and is defined as

@@

In this section, we will assume a network of connectance \(\rho\),
\emph{i.e.} having \(\rho S^2\) interactions (where \(S\) is the species
richness), and \((1-\rho) S^2\) non-interactions. Therefore, the vector
describing the \emph{true} state of the network (assumed to be an
unweighted, directed network) is a column vector
\(\mathbf{o}^T = [\rho, \qquad (1-\rho)]\) (we can safely drop the
\(S^2\) terms, as we will work on the confusion matrix, which ends up
expressing \emph{relative} values). We will apply skill and bias to this
matrix, and measure how a selection of performance metrics respond to
changes in these values, in order to assess their suitability for model
evaluation.

@@

When \(s=0\), \(\text{Tr}(\mathbf{M}) = 0\) (the classifier is
\emph{always} wrong), when \(s=0.5\), the classifier is no-skill and
guesses at random, and when \(s=1\), the classifier is perfect.

@@

In all further simulations, the confusion matrix \(\mathbf{C}\) is
transformed so that it sums to unity, \emph{i.e.} the entries are the
\emph{proportions} of guesses.

@@

In this section, we will change the values of \(b\), \(s\), and
\(\rho\), and report how the main measures discussed in the introduction
(MCC, \(F_1\), \(\kappa\), and informedness) respond. Before we do so,
it is important to explain why we will not focus on accuracy too much.
Accuracy is the number of correct predictions
(\(\text{Tr}(\mathbf{C})\)) divided by the sum of the confusion matrix.
For a no-skill, no-bias classifier, accuracy is equal to
\(\rho^2 + (1-\rho)^2\); for \(\rho = 0.05\), this is \(\approx 0.90\),
and for \(\rho = 0.01\), this is equal to \(\approx 0.98\). In other
words, the values of accuracy are high enough to be uninformative (for
\(\rho\) small, \(\rho^2 \ll (1-\rho)^2\)). More concerning is the fact
that introducing bias changes the response of accuracy in unexpected
ways. Assuming a no-skill classifier, the numerator of accuracy becomes
\(b\rho^2 + (1-b)(1-\rho)^2\), which increases when \(b\) is low, which
specifically means that at equal skill, a classifier that under-predicts
interactions will have higher accuracy than an un-biased classifier
(because the value of accuracy is dominated by the size of tn, which
will increase). These issues are absent from balanced accuracy, but
should nevertheless lead us to not report accuracy as the primary
measure of network prediction success; moving forward, we will focus on
other measures.

In order to examine how MCC, \(F_1\), \(\kappa\), and informedness
change w.r.t. the imbalance, skill, and bias, we performed a grid
exploration of the values of \(\text{logit}(s)\) and \(\text{logit}(b)\)
linearly from \(-10\) to \(10\); \(\text{logit}(x) = -10\) means that
\(x\) is essentially 0, and \(\text{logit}(x) = 10\) means it is
essentially 1 -- this choice was motivated by the fact that most
responses are non-linear with regards to bias and skill. The values or
\(\rho\) were taken linearly in \(]0, 0.5]\), which is within the range
of connectance for species interaction networks. Note that at this
point, there is no network model to speak of; the confusion matrix we
discuss can be obtained for any classification task. Based on the
previous discussion, the desirable properties for a measure of
classifier success should be: an increase with classifier skill,
especially at low bias; a hump-shaped response to bias, especially at
high skill, and ideally centered around \(\text{logit}(b)=0\); an
increase with prevalence up until equiprevalence is reached.

@@

In the following section, we will generate random bipartite networks
(this works without loss of generality on unipartite networks), and
train four binary classifiers (as well as an ensemble model using the
sum of ranged outputs from the component models) on 30\% of the
interaction data. In practice, testing usually uses 70\% of the total
data; for ecological networks, where interactions are sparse \emph{and}
the number of species is low, this may not be the best solution, as the
testing set becomes constrained not by the \emph{proportion} of
interactions, but by their \emph{number}. Preliminary experiments using
different splits revealed no qualitative change in the results. Networks
are generated by picking a random infectiousness trait \(v_i\) for 100
species (from a beta distribution \(B(\alpha=6,\beta=8)\) distribution),
and a resistance trait \(h_j\) for 100 species (from
\(B(\alpha=2,\beta=8)\) distribution). There is an interaction between
\(i\) and \(j\) when \(v_i-\xi/2 \le h_j \le v_i+\xi/2\), where \(\xi\)
is a constant regulating the connectance of the network (there is an
almost 1:1 relationship between \(\xi\) and connectance), and varies
uniformly in \([0.05, 0.35]\). This model gives fully interval networks
that are close analogues to the bacteria--phage model of Weitz et al.
(2005), with both a modular structure and a non-uniform degree
distribution. This dataset is easy for almost any algorithm to learn:
when trained with features \([v_i, h_j, \text{abs}(v_i, h_j)] ^T\) to
predict the interactions between \(i\) and \(j\), all four models
presented below were able to reach almost perfect predictions all the
time (data not presented here) -- this is in part because the rule
(there is maximum value of the distance between traits for which there
is an interaction) is fixed for all interactions. In order to make the
problem more difficult to solve, we use \([v_i, h_j]\) as a feature
vector (\emph{i.e.} the traits on which the models are trained), and
therefore the models will have to uncover that the rule for interaction
is \(\text{abs}(v_i, h_j) \le \xi\). The models therefore all have the
following form, where \(i_{i,j}\) is an interaction from species \(i\)
to species \(j\):

\[
\begin{bmatrix}
    i_{1,1} \\
    i_{1,2} \\
    \vdots \\
    i_{m,n-1} \\
    i_{m,n}
\end{bmatrix}
\propto
\begin{bmatrix}
           v_{1} & h_{1} \\
           v_{1} & h_{2} \\
           \vdots & \vdots \\
           v_{m} & h_{n-1} \\
           v_{m} & h_{n}
         \end{bmatrix}
\]

@@

TODO informedness and accuracy should be easy to beat, make sure the
model is better than them!
